---
title: "Applied example for Poisson Regression"
author: "Irene Alfarone"
date: "2025-05-06"
format:
  html:
    toc: true
    code-fold: true
    code-tools: true
    df-print: paged
execute:
  echo: true
  warning: false
  message: false
---

## Introduction

### Poisson regression

We use a  Poisson distribution when modeling count data, in particular events that 
are rare, independent, and occur over a fixed observation window (time, space, etc.).

Some key characteristics:

* **Discrete Events**: The number of events is a count (0, 1, 2, ...)
* **Lower Bound**: Negative counts are not possible (i.e., we do not have -1 events ;))
* **Randomness**: Events occur independently and randomly
* **Constant Rate**: Events happen at a constant average rate over time or space
* **No Simultaneity**: Two or more events do not occur at the exact same time.

* The **expected value** (mean) is:  
  $$
  \mathbb{E}[Y] = \lambda
  $$

* The **variance** is also:  
  $$
  \text{Var}(Y) = \lambda
  $$

This means that in the Poisson distribution, the **mean and the variance are equal**.

In Poisson regression, we model the mean count $\lambda$ as a function of predictors using the log link:

$$
\log(\lambda) = \beta_0 + \beta_1 X_1 + \dots + \beta_k X_k
$$

To interpret the coefficients, we exponentiate the values of $\beta$:  
  $$
  \exp(\beta_j)
  $$

This gives a **rate ratio**: it tells us **how many times higher (or lower)** the expected number of events is for a one-unit increase in the predictor $X_j$, holding all else constant.


### Simulated example
We simulate a dataset in which the number of anxiety attacks experienced by individuals over a four-week period are reported. The primary objective is to model the count data using Poisson regression, considering two predictors:

- **Therapy**: patient randomized to special CBT treatment or Std treatment (`CBT` or `Std`).
- **Stress**: continuous outcome variable, representing perceived stress.

We aim to assess how therapy and stress levels influence the frequency of anxiety attacks.

## Data Simulation
```{r}
library(ggplot2)
library(dplyr)
library(MASS)

set.seed(2025)  
N = 200
Therapy = rbinom(N, 1, 0.5)
Stress = runif(N, min = 1, max = 5)

lambda = exp(0.2 + 0.5 * Stress - 0.8 * Therapy)

Attacks = rpois(N, lambda)

d = data.frame(
  Attacks = Attacks,
  Therapy = factor(Therapy, labels = c("Std", "CBT")),
  Stress = round(Stress, 0)
)

head(d)
```

In a Poisson regression model, the outcome variable $Y$ (e.g., number of anxiety attacks) is assumed to follow a Poisson distribution:

$$
Y \sim \text{Poisson}(\lambda)
$$

where $\lambda$ is the expected count (i.e., $\mathbb{E}[Y]$), and must be strictly positive.

### The Log Link Function

Poisson regression uses the logarithmic link function to relate the predictors to the mean of the distribution:

$$
\log(\lambda) = \beta_0 + \beta_1 \cdot \text{Stress} + \beta_2 \cdot \text{Therapy}
$$

This ensures that $\lambda > 0$ for all values of the predictors, which is a key requirement for count data. The exponential transformation gives us the expected number of events:

$$
\lambda = \exp(\beta_0 + \beta_1 \cdot \text{Stress} + \beta_2 \cdot \text{Therapy})
$$


```{r}
ggplot(d, aes(x = Stress, y = Attacks, color = Therapy)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "glm", method.args = list(family = "poisson"), se = FALSE) +
  labs(
    title = "Anxiety Attacks by Stress and Therapy",
    x = "Stress",
    y = "Number of Anxiety Attacks",
    color = "Therapy"
  ) +
  theme_minimal()
```
## The model

The simulated data model the number of anxiety attacks experienced by individuals based on their stress and their therapy condition. The number of attacks is a **count variable**, therefore we use Poisson regression!


We fit a generalized linear model (GLM) with the **Poisson family** and a **log link** function, assuming the following form:

$$
\log(\mathrm{Attacks}_i) = \beta_0 + \beta_1 \cdot \text{Stress}_i + \beta_2 \cdot \text{Therapy}_i
$$

Here:

* $\mathrm{Attacks}_i$ is the expected number of anxiety attacks for individual $i$,

* `Stress` is a continuous predictor,

* `Therapy` is a binary categorical predictor (0 = Std, 1 = CBT).


```{r}
m = glm(Attacks ~ Therapy + Stress, family = poisson, data = d)
summary(m)
exp(coef(m))
```
```{r}
mnb = glm.nb(Attacks ~ Therapy + Stress, data = d)
summary(mnb)
exp(coef(mnb))
```

### Interpretation of Coefficients

* **Intercept ($\beta_0$)**: Represents the log-mean number of anxiety attacks when both predictors are at zero. 

* **Stress ($\beta_1$)**: A positive value indicates that as perceived stress increases. For example, if $\beta_1 = 0.5$, then a one-unit increase in stress corresponds to a $e^{0.5} \approx 1.65$ increase in expected attacks. 

* **Therapy ($\beta_2$)**: A negative coefficient suggests that therapy is associated with fewer anxiety attacks, holding stress constant. If $\beta_2 = -0.8$, the special CBT reduces the expected attack rate by about $e^{-0.8} \approx 0.45$, or a 55% reduction.


### Statistics mumbo-jumbo

* **Why Poisson?**  
Counts like anxiety episodes are non-negative integers with potentially skewed distributions. Using linear regression would violate assumptions of normality and homoscedasticity and might predict impossible negative values.

* **Log-link intuition**  
The log link allows the model to predict only positive expected counts. 

* **Overdispersion**  
It is important to check for overdispersion (i.e., when variance > mean), which is common in psychological data. If present, a negative binomial model might be more appropriate.

### TRUE and FALSE

* If the coefficient for therapy is negative, then receiving therapy decreases the expected number of anxiety attacks.

* The Poisson model assumes that the variance of the outcome is always greater than the mean.

* If the model shows overdispersion (variance > mean), a negative binomial model may be more appropriate.

* In Poisson regression, the expected count is modeled as a log-linear function of the predictors.

* The exponentiated coefficients in a Poisson regression model represent rate ratios.

* It is appropriate to include both continuous and categorical predictors in a Poisson regression model.


